---
title: "Comparison with unmarked package"
output: pdf_document
# output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Comparison with unmarked package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(roccupy)
```

The goal of this vignette is to show how one can convert from the data format used in `unmarked` to the one used in `roccupy`, and that the maximum likelihood algorithm in `roccupy` closely matches the estimates from `occu` in `unmarked`.

First, load the example data in the `unmarked` package:

```{r}
library(unmarked)
wt <- read.csv(system.file("csv","widewt.csv", package="unmarked"))
```
The data looks as follows:

```{r}
head(wt)
```
While `unmarked` uses what one might call a "wide" data format, `roccupy` uses a "long" format. We will have to melt the observation covariates. To do so, let us first pull out the columns that belong together:

```{r}
site_covs <- wt[, c('elev', 'forest', 'length')]

obs_cov_1 <- wt[, c('site', 'date.1', 'date.2', 'date.3')]
obs_cov_2 <- wt[, c('site', 'ivel.1', 'ivel.2', 'ivel.3')]

y <- wt[, c('site', 'y.1', 'y.2', 'y.3')]

head(obs_cov_1)
```
Now, we _melt_ each of them so that each row in the result corresponds to a single visit.

```{r}
library(reshape2)

# Melt date covariate
melted_cov_1 <- melt(obs_cov_1, id.vars='site')
sites_1 <- melted_cov_1$site
vals_1 <- melted_cov_1$value

# Melt ivel covariate
melted_cov_2 <- melt(obs_cov_2, id.vars='site')
sites_2 <- melted_cov_2$site
vals_2 <- melted_cov_2$value

# Melt presence/absence
melted_y <- melt(y, id.vars='site')
sites_y <- melted_y$site
vals_y <- melted_y$value

# Double-check that the order appears unchanged:
all(sites_1 == sites_2)
all(sites_2 == sites_y)
```
The key motivation for using the long format is that it is _sparse_. Rather than having to fill missing visits with `NA`, they are simply omitted in the long format. We now put together all the observation-level information and discard the `NA` rows. Note that we subtract 1 from each site so that numbering is _zero-based_ -- this is because `roccupy` uses python under the hood, which uses zero-based indexing.

```{r}
cur_data <- data.frame(cbind(date=vals_1, ivel=vals_2, site=sites_1 - 1, y=vals_y))

no_nas <- na.omit(cur_data)

# Check how many were removed
print(dim(cur_data))
print(dim(no_nas))
```
In this case, the savings are negligible. For citizen science data like eBird, however, they are considerable.

Now, we can pull out the arrays we need for the `roccupy` model:

```{r}
X_obs <- no_nas[, c('date', 'ivel')]
checklist_cell_ids <- no_nas$site
X_env <- site_covs
y_checklist <- data.frame(cur_species=no_nas$y)
```

Now, fit the model:

```{r}
result <- ssod_ml('forest + elev + length', 'date + ivel', X_env, X_obs, 
                  y_checklist, checklist_cell_ids)
```

We can extract the estimated coefficients:

```{r}
coef(result)
```

And we can compare this with what `unmarked` produces:

```{r}
y <- wt[ ,2:4]

siteCovs <-  wt[ ,5:7]

obsCovs <- list(date = wt[ ,8:10],
                ivel = wt[ ,11:13])

umf <- unmarkedFrameOccu(y = y, siteCovs = siteCovs, obsCovs = obsCovs)

fm1 <- occu(formula = ~ date + ivel
                      ~ forest + elev + length,
            data = umf)
```

Agreement is very close:

```{r}
print(coef(fm1, 'state'))
print(coef(fm1, 'det'))
```
Prediction works as follows:

```{r}
# Predict suitability:
suit_pred <- predict(result, X_env, type='env')

# Predict probability of observing:
# Note that here X_env and X_obs lengths must match.
obs_pred <- predict(result, X_env[checklist_cell_ids + 1, ], X_obs, type='obs')

head(obs_pred)
```