---
title: "Using roccupy with data from auk"
# output: rmarkdown::html_vignette
output: pdf_document
vignette: >
  %\VignetteIndexEntry{Using roccupy with data from auk}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(roccupy)
library(auk)
```

The goal of this vignette is to illustrate how data can be converted to the format used in `roccupy`. In particular, we'll be looking at how to convert data from the [`auk` package](https://cornelllabofornithology.github.io/auk/index.html) to the required format. Although we will be using only a very small portion of eBird data -- the amount that is included with `auk` -- the steps should be the same for much larger analyses using eBird.

First, we'll follow the steps from the [`auk` introduction](https://cornelllabofornithology.github.io/auk/articles/auk.html) to obtain a small sample of zero-filled records from eBird. If you'd like to know more about what these steps are doing, please read the introduction linked.

```{r}
# to produce zero-filled data, provide an EBD and sampling event data file
f_ebd <- system.file("extdata/zerofill-ex_ebd.txt", package = "auk")
f_smp <- system.file("extdata/zerofill-ex_sampling.txt", package = "auk")
filters <- auk_ebd(f_ebd, file_sampling = f_smp) %>% 
  auk_complete()
filters
```
```{r}
ebd_sed_filtered <- auk_filter(filters, 
                               file = "ebd-filtered.txt",
                               file_sampling = "sampling-filtered.txt",
                               overwrite = TRUE)
ebd_sed_filtered

ebd_zf <- auk_zerofill(ebd_sed_filtered)
```

The zero-filled data consists of two tables. One lists the species observations:

```{r}
str(ebd_zf$observations)
```
The other lists the details of each checklist, such as the time spent looking:

```{r}
str(ebd_zf$sampling_events)
```

First, we'll reshape the observations into the format required for `roccupy`. What we need is a DataFrame of shape `(n_checklists, n_species)`, rather than the long format produced by `auk`. Luckily, we can easily switch to the required format using the `reshape2` package:

```{r}
library(reshape2)

y_checklist <- dcast(ebd_zf$observations, checklist_id ~ scientific_name, 
                     value.var = 'species_observed')

species_names <- unique(ebd_zf$observations$scientific_name)

str(y_checklist)
```

In the following, we will have to drop some records since they are missing information. To make this a bit easier, we'll put together the observations and checklist details into a single DataFrame:

```{r}
combined <- merge(ebd_zf$sampling_events, y_checklist, by='checklist_id')
```

Now, we need to find some environmental covariates and define our sites. Here, we'll use the `worldclim` variables to do both: the covariates in `worldclim` will be our environmental covariates, and the cells in the raster will be our sites. We emphasise here that this example is _not_ meant to be ecologically meaningful: the amount of data is small and it is collected in a very small area, so it is unlikely that the inferences are very insightful.

```{r}
library(raster)

r <- getData("worldclim",var="bio",res=2.5)
```

Let's fetch the cell numbers, which will be our sites:

```{r}
cell_nums <- cellFromXY(r, cbind(combined$longitude, combined$latitude))
table(cell_nums)
```
We have some repeat observations, which is good. To keep track of the cell numbers, add them to the sampling information:

```{r}
combined$env_cell_num <- cell_nums
```

Now we'll fetch the environmental covariates in these cells:

```{r}
unique_cells <- unique(cell_nums)
cell_vals <- raster::extract(r, unique_cells)

X_env <- data.frame(cell_vals)
```

Some of them are NA. We will drop these. We will also drop checklists without durations.

```{r}
X_env$cell_num <- unique_cells
na_cells <- rowSums(is.na(X_env)) > 0
to_drop <- unique_cells[na_cells]

# Drop NA cells from sampling info:
rel_combined <- combined[!(combined$env_cell_num %in% to_drop), ]

# Also only keep those with durations recorded
rel_combined <- rel_combined[!is.na(rel_combined$duration_minutes), ]

# Drop the NA environmental cells
rel_X_env <- X_env[!(X_env$cell_num %in% to_drop), ]
```

The next step is key: we need to match up the checklists with the corresponding sites. We can use R's `match` function to make this easy. However, crucially, sites are numbered from _zero_ onwards, so we need to subtract 1 from the result of `match`:

```{r}
# Now, we need to encode the cell ids
cell_ids <- match(rel_combined$env_cell_num, rel_X_env$cell_num)

# IMPORTANT: These need to start from zero! So:
cell_ids <- cell_ids - 1
```

Now we're pretty much good to go! Let's fit an example model with three environmental covariates and log-transformed duration:

```{r}
env_vars_to_use <- c('bio1', 'bio2', 'bio4')

# Scale the environmental variables:
to_scale <- rel_X_env[, env_vars_to_use]
scaled_X_env <- scale(to_scale)
scaled_X_env_df <- data.frame(scaled_X_env)

env_formula <- paste(env_vars_to_use, collapse='+')

env_formula
```

```{r}
X_obs <- data.frame(log_duration = log(rel_combined$duration_minutes))

obs_formula <- 'log_duration'
```

Now we have all we need and can fit the model:

```{r}
checklist_cell_ids <- cell_ids

y_checklist <- rel_combined[, species_names]

fit_result <- msod_vi(env_formula, obs_formula, scaled_X_env_df, X_obs, 
                      y_checklist, checklist_cell_ids)
```

This object can now be used in the usual way. For example, we can take a look at the draws for the observation part of the model:

```{r}
coef_draws <- coef(fit_result)
lapply(coef_draws$obs_slopes, summary)
```

Please see the `ebird-example` vignette for more detail about how to interpret the results of the model.